### 2013-11-12 src2/embarbb/programming.tex
hacking
update

%The vectors 
%in ArBB can be one, two or three dimensional dense vectors (in other words arrays)
%or nested vectors that allow the programmer to utilise nested data parallelism. 
%A nested vector is represented by an array of elements and an array holding a 
%nesting descriptor. 
% BJS: I have seen this elsewhere but cannot find it! (wanted to \cite something)
%The ArBB system offers the programmer {\em dense} and {\em nested} vectors 
%and operations on these. A dense vector (or simply put, an array) can 
%be one, two or three dimensional. A nested vector is represented by an 
%array of elements and an array holding a nesting descriptor. Nested vectors 
%can be used to express nested data-parallelism.

@@

After setting up and copying the data into ArBB, the function {\tt arbb\_matrix\_vector} 
is {\em called} using the {\tt call} command. At this point, a number of
things happen. If the function is called for the first time, it is 
{\em captured}. This means that the function is run as a C++ function, which produces 
an intermediate representation (IR) of the computation (which is standard in such embeddings).
For example, if the function being captured contains a normal C++ {\em for loop}, it will be unrolled, much in the same way as recursion in a
Haskell embedded DSL results in unrolled code. After the function is captured, 
the IR is further optimised and finally executed. Compilation and capture 
only occur the first time a function is called. The compilation procedure
is outlined in~\cite{ARBB2011}.

@@

% \FloatBarrier 
\subsubsection{Saxpy} 
{\em Saxpy} is a vector operation that gets its name from the 
description of what it does ``{\em S}ingle-precision {\em A}lpha {\em X} {\em P}lus {\em Y}''. 
There is also a double-precision version called {\em daxpy}. Here we can use a single source 
to obtain both versions:
%. The code listing below {\em axpy} illustrates this:

\begin{verbatim} 
saxpy :: Num a 
      => Exp a 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
saxpy s x y = (ss*x) + y 
    where 
      ss = constVector (length x) s
\end{verbatim}

@@

\begin{verbatim}
main = 
  withArBB $ 
  do 
     f <- capture saxpy

@@

     liftIO$ putStrLn$ show r
\end{verbatim}

@@

\begin{verbatim}
main =  
  withArBB $ 
  do 
     f <- capture matVec  
     let m1 = V.fromList [2,0,0,0,
                          0,2,0,0,
                          0,0,2,0,
                          0,0,0,2]
         v1 = V.fromList [1,2,3,4]

@@

     liftIO$ putStrLn$ show r
\end{verbatim}
%$ -- fixes syntax highlighting
The complete Haskell version of this program is slightly shorter than the corresponding 
C++ version. However, the C++ version could probably be made shorter as well; the Haskell and C++ versions must be considered similar in implementation complexity.

@@

\begin{verbatim}
histImage :: Exp (Vector Word32) 
           -> Exp (DVector Dim2 Word8) 
histImage input = fst $ while cond body (cvn,0)    
    where 
      cond (img,i) = i <* n 
      body (img,i) = (replaceCol i col' img,i+1) 
          where 
            val = input ! i 
            col = extractCol i img
            col' = fill black 0 n col
            n = 255 - scale 255 m val

      n = length input 
      cv = constVector (n*n) white  
      cvn = setRegularNesting2D n n cv
      m   = index0 (maxReduce rows input)
      black = 0 
      white = 255
\end{verbatim}

@@

The multiplication acts pointwise on the elements of {\tt vals}
and the vector (of the same length) containing relevant elements of the dense vector (produced using {\tt gather1D}).
The function {\tt applyNesting} produces
a nested vector of rows. The function {\tt addReduceSeg} sums each
of the rows (or segments) in that array, producing the non-zero elements
of the result vector. Note that there is irregular data parallelism
here because the rows may have different lengths. We would hope
to get parallelism both in the individual summations and between summations.
This is a well known algorithm that can be traced back to Blelloch~\cite{NESL}.

>>>>

%The vectors 
%in ArBB can be one, two or three dimensional dense vectors (in other words arrays)
%or nested vectors that allow the programmer to utilise nested data parallelism. 
%A nested vector is represented by an array of elements and an array holding a 
%nesting descriptor. 
% BJS: I have seen this elsewhere but cannot find it! (wanted to \citeemb something)
%The ArBB system offers the programmer {\em dense} and {\em nested} vectors 
%and operations on these. A dense vector (or simply put, an array) can 
%be one, two or three dimensional. A nested vector is represented by an 
%array of elements and an array holding a nesting descriptor. Nested vectors 
%can be used to express nested data-parallelism.

@@

After setting up and copying the data into ArBB, the function {\tt arbb\_matrix\_vector} 
is {\em called} using the {\tt call} command. At this point, a number of
things happen. If the function is called for the first time, it is 
{\em captured}. This means that the function is run as a C++ function, which produces 
an intermediate representation (IR) of the computation (which is standard in such embeddings).
For example, if the function being captured contains a normal C++ {\em for loop}, it will be unrolled, much in the same way as recursion in a
Haskell embedded DSL results in unrolled code. After the function is captured, 
the IR is further optimised and finally executed. Compilation and capture 
only occur the first time a function is called. The compilation procedure
is outlined in~\citeemb{ARBB2011}.

@@

\FloatBarrier 
\subsubsection{Saxpy} 
{\em Saxpy} is a vector operation that gets its name from the 
description of what it does ``{\em S}ingle-precision {\em A}lpha {\em X} {\em P}lus {\em Y}''. 
There is also a double-precision version called {\em daxpy}. Here we can use a single source 
to obtain both versions:
%. The code listing below {\em axpy} illustrates this:

\begin{Verbatim}[samepage=true]
saxpy :: Num a 
      => Exp a 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
saxpy s x y = (ss*x) + y 
    where 
      ss = constVector (length x) s
\end{Verbatim}

@@

\begin{Verbatim}[samepage=true]
main = withArBB $ 
  do 
     f <- capture saxpy

@@

     liftIO$ putStrLn$ show r
\end{Verbatim}

% $

@@

\begin{Verbatim}[samepage=true]
main = withArBB $ 
  do 
     f <- capture matVec  
     let m1 = V.fromList [2,0,0,0,
                          0,2,0,0,
                          0,0,2,0,
                          0,0,0,2]
         v1 = V.fromList [1,2,3,4]

@@

     liftIO$ putStrLn$ show r
\end{Verbatim}
%$ -- fixes syntax highlighting
The complete Haskell version of this program is slightly shorter than the corresponding 
C++ version. However, the C++ version could probably be made shorter as well; the Haskell and C++ versions must be considered similar in implementation complexity.

@@

\begin{Verbatim}[samepage=true]
histImage :: Exp (Vector Word32) 
           -> Exp (DVector Dim2 Word8) 
histImage input = fst $ while cond body (cvn,0)    
    where 
      cond (img,i) = i <* n 
      body (img,i) = (replaceCol i col' img,i+1) 
          where 
            val = input ! i 
            col = extractCol i img
            col' = fill black 0 n col
            n = 255 - scale 255 m val

      n = length input 
      cv = constVector (n*n) white  
      cvn = setRegularNesting2D n n cv
      m   = index0 (maxReduce rows input)
      black = 0 
      white = 255
\end{Verbatim}

% $

@@

The multiplication acts pointwise on the elements of {\tt vals}
and the vector (of the same length) containing relevant elements of the dense vector (produced using {\tt gather1D}).
The function {\tt applyNesting} produces
a nested vector of rows. The function {\tt addReduceSeg} sums each
of the rows (or segments) in that array, producing the non-zero elements
of the result vector. Note that there is irregular data parallelism
here because the rows may have different lengths. We would hope
to get parallelism both in the individual summations and between summations.
This is a well known algorithm that can be traced back to Blelloch~\citeemb{NESL}.

### 2013-11-12 src2/embarbb/results.tex
update

\subsubsection{Sobel edge detection}
In the Repa Sobel code, only the applications of the stencils are timed. This 
part is defined as follows 
\begin{verbatim}
gradientX :: Monad m => Image -> m Image
gradientX img
        = computeP
        $ forStencil2 (BoundConst 0) img
              [stencil2| -1  0  1
                         -2  0  2
                         -1  0  1 |]

gradientY :: Monad m => Image -> m Image
gradientY img
        = computeP
        $ forStencil2 (BoundConst 0) img
             [stencil2|  1  2  1
                         0  0  0
                        -1 -2 -1 |] 
\end{verbatim}

>>>>

\subsubsection{Sobel edge detection}
In the Repa Sobel code, only the applications of the stencils are timed. This 
part is defined as follows 
\begin{Verbatim}[samepage=true]
gradientX :: Monad m => Image -> m Image
gradientX img
        = computeP
        $ forStencil2 (BoundConst 0) img
              [stencil2| -1  0  1
                         -2  0  2
                         -1  0  1 |]
\end{Verbatim}
\begin{Verbatim}[samepage=true]
gradientY :: Monad m => Image -> m Image
gradientY img
        = computeP
        $ forStencil2 (BoundConst 0) img
             [stencil2|  1  2  1
                         0  0  0
                        -1 -2 -1 |] 
\end{Verbatim}

### 2013-11-12 src2/ifl/related.tex
update

The more general purpose languages aim to abstract away from the graphics heritage of GPUs, and target a larger group of programmers.
The thesis by Lucas presents CGiS, an imperative data-parallel programming
language that targets both GPUs and SIMD capable CPUs -- with the aim being
a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\citeifl{Lucas08}.
BrookGPU (which is usually just called Brook) is a classic example of a language~\citeifl{Brook} designed to raise the level of abstraction at which GPGPU programming is done. It is an extension of C with embedded kernels, aimed at arithmetic-intense data parallel computations. C is used to declare streams\footnote{Brook is referred to as a ``stream processing'' language, but this means something different from what the reader might expect: a stream in this context is
a possibly multi-dimensional array of elements, each of which can be processed separately, in parallel.}, CG/HLSL (the lower level GPU languages) to declare kernels, while
function calls to a runtime library direct the execution of the program.
Brook had significant impact in that it raised the level of abstraction at which GPGPU programming can be done. The language Sh also aimed to raise the level of abstraction at which GPUs were programmed~\citeifl{Sh}. Sh was an embedded language in C++, so our work is close in spirit to it. Sh has since evolved into the RapidMind development platform~\citeifl{RapidMind}, which now supports multicores and Cell processors as well as GPUs. 
The RapidMind programming model has arrays as first class types. It
has been influenced by functional languages like NESL and SETL, and its program objects are pure functions. Thus it supports both functional and imperative styles of programming.
A recent PhD thesis by Jansen asserts that there are some problems with RapidMind's use of macros to embed the GPU programming language in C++, including the inability to pass
kernels (or shader programs) as classes~\citeifl{JansenThesis}; the thesis proposes GPU++ and claims improvement over previous approaches, particularly through the exploitation of automatic partitioning of the programs onto the available GPU hardware, and through compiler optimisations that improve runtime performance.

>>>>

The more general purpose languages aim to abstract away from the graphics heritage of GPUs, and target a larger group of programmers.
The thesis by Lucas presents CGiS, an imperative data-parallel programming
language that targets both GPUs and SIMD capable CPUs -- with the aim being
a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\citeifl{Lucas08}.
BrookGPU (which is usually just called Brook) is an example of a language~\citeifl{Brook} designed to raise the level of abstraction at which GPGPU programming is done. It is an extension of C with embedded kernels, aimed at arithmetic-intense data parallel computations. C is used to declare streams\footnote{Brook is referred to as a ``stream processing'' language, but this means something different from what the reader might expect: a stream in this context is
a possibly multi-dimensional array of elements, each of which can be processed separately, in parallel.}, CG/HLSL (the lower level GPU languages) to declare kernels, while
function calls to a runtime library direct the execution of the program.
Brook had significant impact in that it raised the level of abstraction at which GPGPU programming can be done. The language Sh also aimed to raise the level of abstraction at which GPUs were programmed~\citeifl{Sh}. Sh was an embedded language in C++, so our work is close in spirit to it. Sh has since evolved into the RapidMind development platform~\citeifl{RapidMind}, which now supports multicores and Cell processors as well as GPUs. 
The RapidMind programming model has arrays as first class types. It
has been influenced by functional languages like NESL and SETL, and its program objects are pure functions. Thus it supports both functional and imperative styles of programming.
A recent PhD thesis by Jansen asserts that there are some problems with RapidMind's use of macros to embed the GPU programming language in C++, including the inability to pass
kernels (or shader programs) as classes~\citeifl{JansenThesis}; the thesis proposes GPU++ and claims improvement over previous approaches, particularly through the exploitation of automatic partitioning of the programs onto the available GPU hardware, and through compiler optimisations that improve runtime performance.

### 2013-11-12 src2/thesis.tex
update

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. There are many other benefits of embedded languages: making use 
of the host language's library functions (if they are general enough), the host language 
can be used as a powerful macro language for the embedded language, providing a more 
leveled learning curve to programmers already familiar with the host language. 
For programmers who are already proficient in the host language, learning an embedded 
language is more like picking up a new library.

@@

The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
specified above. Each input array passed to this function must have length a multiple of 32.
Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
used to map the local computation onto each chunk of the input arrays. The result is an 
array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
is generated from the {\tt vecAddG} program:

@@

Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
a reduction kernel in Obsidian. Seven reduction kernels are implemented, the first 
very simple and naive. Each following kernel tweaks some aspect of the previous one. 
Applying these changes is pleasingly simple using Obsidian; some of these changes 
would mean a major rewrite in CUDA. The final kernel is almost 5 times as fast as the first, 
and reaches a throughput of 140GB/s on a GPU with 192GB/s memory bandwidth. The speed of any 
kernel will always be limited by how fast the data can be pushed through the GPU; this performance 
is quite satisfactory. In Paper F, a large reduction algorithm based on this kernel is 
implemented that runs slightly faster than Accelerates fold skeleton. With this reduction 
case study, we have pushed Obsidian as far as possible with the current abstractions. 
However, it is possible that with more work on the warp level abstractions, this could be 
improved further. Our reduction case study was based on lecture slides from 
NVIDIA~\citet{reducion}. Reference~\citet{merrill}, has much information about 
how to think about performance when programming GPUs. 
%faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
%reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
%been happy just getting close.

@@

ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
\begin{itemize} 
\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting:}      {\tt sort} 
\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil:} are all supported via a single primitive called {\tt map}.
\end{itemize}

These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential looping 
constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just adds a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell embedded languages, but it may be even 
more surprising to a C++ programmer. Before an ArBB function can be computed, it needs to be 
captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
functionality. The first time {\tt call} is used on a function pointer, that function is 
executed and the ArBB functions used in it build an AST (we say the function has been 
captured). The AST is then compiled and specialised for the hardware available. Any 
subsequent calls of a function using ArBB functionality do not lead to a recompilation; 
as the first call caches this compiled code.

>>>>

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. There are many other benefits of embedded languages: making use 
of the host language's library functions (if they are general enough), the host language 
can be used as a powerful macro language for the embedded language, providing a more 
levelled learning curve to programmers already familiar with the host language. 
For programmers who are already proficient in the host language, learning an embedded 
language is more like picking up a new library.

@@

The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
specified above. Each input array passed to this function must have length a multiple of 32.
Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
used to map the local computation onto each chunk of the input arrays. The result is an 
array of arrays that is flattened with the {\tt pConcat} function.

\noindent The CUDA code below is generated from the {\tt vecAddG} program:

@@

Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
a reduction kernel in Obsidian. Seven reduction kernels are implemented, the first 
very simple and naive. Each following kernel tweaks some aspect of the previous one. 
Applying these changes is pleasingly simple using Obsidian; some of these changes 
would mean a major rewrite in CUDA. The final kernel is almost 5 times as fast as the first, 
and reaches a throughput of 140GB/s on a GPU with 192GB/s memory bandwidth. The speed of any 
kernel will always be limited by how fast the data can be pushed through the GPU
(see reference~\citet{merrill} for an in depth discussion of kernel performance); this performance 
is very satisfactory. In Paper F, a large reduction algorithm based on this kernel is 
implemented that runs slightly faster than Accelerate's fold skeleton. Even being just nearly 
as fast as Accelerate would in this case have been fine, since the fold skeleton is hand-tuned 
CUDA code. With this reduction case study, we have pushed Obsidian as far as possible 
with the current abstractions. However, it is possible that with more work on the warp 
level abstractions, this could be improved further. Our reduction case study was based 
on lecture slides from NVIDIA~\citet{reduction}.
%faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
%reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
%been happy just getting close.

@@

ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
\begin{small}
\begin{itemize} 
\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting:}      {\tt sort} 
\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil:} are all supported via a single primitive called {\tt map}.
\end{itemize}
\end{small}

\noindent These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential looping 
constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just adds a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell embedded languages, but it may be even 
more surprising to a C++ programmer. Before an ArBB function can be computed, it needs to be 
captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
functionality. The first time {\tt call} is used on a function pointer, that function is 
executed and the ArBB functions used in it build an AST (we say the function has been 
captured). The AST is then compiled and specialised for the hardware available. Any 
subsequent calls of a function using ArBB functionality do not lead to a recompilation; 
as the first call caches this compiled code.
